Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-07 method.tex
lloid_paper: Forgot {}, so a macro ate a space
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@959 ccdd05d5-56c1-461a-bf98-85f6876523b4
lloid_paper: Changes from Mireia and misc polishing and clipping
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@961 ccdd05d5-56c1-461a-bf98-85f6876523b4

Here we explore a method for reducing the computational cost of a \TD\ search
for compact binary coalescence.  We will give a truly zero-latency algorithm
that competes in terms of floating point operations per second with the
conventional \FD\ method, which requires a signficant latency in order to be
computationally cheap. Our method, \lloid\footnote{\lloid stands for \emph{Low
Latency Online Inspiral Detection} and was a code name for this procedure
throughout development. Although the name doesn't carry any specific
information about the algorithm, it became known within the gravitational-wave
community and has stuck.}%
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.

The first transformation is to chop the \TD{} templates up into \emph{time
slices}.  Since each template slice is disjoint in time, the resulting set for
a single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim$100 samples
from the early part of the waveform and allows the filters to be evaluated at
about $\sim$100 times lower rate.  This amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the \SVD{} to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of
$\sim$100.  The combined methods reduce the number of floating point operations
to the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

Taking into account the decimation of the detector data, the orthogonal \fir\ filters, the reconstruction of the time-sliced templates, the interpolation of \SNR\ from previous time slices, and the accumulation of \SNR, in total the \lloid\ algorithm requires
$$
\sum_{\mathclap{s=0}}^{\mathclap{S-1}} \left( 2 \slicessamps \numsvdtmps + 2 \numtmps \numsvdtmps + \numtmps \right) f^s + 2\sum_{\mathclap{f \in \{f^s\}}} \left( N^\shortdownarrow + \numtmps N^\shortuparrow \right) f \textrm{ \flops}\,.
$$
The second sum is carried out over the set of sample rates of all of the time slices rather than over the time slices themselves because some time slices may have the same sample rates.

>>>>

Here we explore a method for reducing the computational cost of a \TD\ search
for compact binary coalescence.  We will give a truly zero-latency algorithm
that competes in terms of floating point operations per second with the
conventional \FD\ method, which requires a signficant latency in order to be
computationally cheap. Our method, \lloid{}\footnote{\lloid{} stands for
\emph{Low
Latency Online Inspiral Detection} and was a code name for this procedure
throughout development. Although the name doesn't carry any specific
information about the algorithm, it became known within the gravitational-wave
community and has stuck.}%
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.

The first transformation is to chop the \TD{} templates up into \emph{time
slices}.  Since each template slice is disjoint in time, the resulting set for
a single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling reduces the number of data samples by a
factor of $\sim$100 from the early part of the waveform and allows the filters
to be evaluated at about $\sim$$1/100^\mathrm{th}$ the sample rate. This
amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the \SVD{} to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of
$\sim$100.  The combined methods reduce the number of floating point operations
to the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

Taking into account the decimation of the detector data, the orthogonal \fir\ filters, the reconstruction of the time-sliced templates, the interpolation of \SNR\ from previous time slices, and the accumulation of \SNR, in total the \lloid\ algorithm requires
$$
\sum_{\mathclap{s=0}}^{\mathclap{S-1}} \left( 2 \slicessamps \numsvdtmps + 2 \numtmps \numsvdtmps + \numtmps \right) f^s + 2\sum_{\mathclap{f \in \{f^s\}}} \left( N^\shortdownarrow + \numtmps N^\shortuparrow \right) f \textrm{ \flops}\,.
$$
The second sum is carried out over the set of distinct sample rates rather than over the time slices themselves.

### 2011-06-07 results.tex
lloid_paper: Changes from Mireia and misc polishing and clipping
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@961 ccdd05d5-56c1-461a-bf98-85f6876523b4

We examine the performance of the \lloid\ algorithm on a small region of
compact binary parameter space centered on typical NS-NS masses.  We begin by
constructing a template bank that spans component masses between
1~and~3~$M_\odot$ using a simulated advanced LIGO noise curve~\cite{lal}.  Then
we create sub banks based on chirp mass slices in that parameter space.
Figure \ref{fig:tmpltbank} illustrates this procedure.
\begin{figure}[htbp]
	\begin{center}
		\includegraphics{figures/tmpltbank.png}
		\caption{\label{fig:tmpltbank}Placement of template parameters used in this paper.  The template bank consists of 98544 templates with component masses $m_1$, $m_2$, between 1~and~3~$M_\odot$.  We shall design a filter bank to search for a small subset of these template parameters with chirp masses $\mathcal M$ between 1.1955~and~1.2045~$M_\odot$.}
	\end{center}
\end{figure}
The result is that we obtain 657 templates with chirpmasses between 1.1955~and~1.2045~$M_\odot$.  With this
sub-bank we were able to construct an efficient time slice decomposition that consisted of 11 time slices
with sample rates between 32 and 4096 Hz shown in table \ref{tab:time_slices}.
\begin{table}
\begin{indented}
\caption{\label{tab:time_slices} Filter design for these 657 templates.  From left to right, this table shows the sample rate, time interval, number of samples, and number of orthogonal templates for each time slice.  The \SVD{} tolerance is varied from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.}
\item[]\begin{tabular}{rr@{,}lc*{6}{r}}
\br
\multicolumn{4}{c}{} &\centre{6}{$\log_{10}$ (1 - \textsc{svd} tolerance)} \\
\ns
\multicolumn{4}{c}{} &\crule{6} \\
$f^s$ (Hz) & $[t^s$&$t^{s+1})$ (s) & $N^s$ & -1 & -2 & -3 & -4 & -5 & -6 \\
\mr
4096 & [0&0.5) & 2048 & 1 & 4 & 6 & 8 & 10 & 14 \\
512 & [0.5&4.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 16 \\
256 & [4.5&12.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 15 \\
128 & [12.5&76.5) & 8192 & 6 & 20 & 25 & 28 & 30 & 32 \\
64 & [76.5&140.5) & 4096 & 1 & 8 & 15 & 18 & 20 & 22 \\
64 & [140.5&268.5) & 8192 & 1 & 7 & 21 & 25 & 28 & 30 \\
64 & [268.5&396.5) & 8192 & 1 & 1 & 15 & 20 & 23 & 25 \\
32 & [396.5&460.5) & 2048 & 1 & 1 & 3 & 9 & 12 & 14 \\
32 & [460.5&588.5) & 4096 & 1 & 1 & 7 & 16 & 18 & 21 \\
32 & [588.5&844.5) & 8192 & 1 & 1 & 8 & 26 & 30 & 33 \\
32 & [844.5&1100.5) & 8192 & 1 & 1 & 1 & 12 & 20 & 23 \\
\br
\end{tabular}
\end{indented}
\end{table}
We use this template bank and decomposition for the remainder of this section.

@@

It is also within reason that some \SNR\ loss could arise from other suboptimal
implementations of the \lloid\ algorithm in our test pipeline.  To gauge
accurately how well the pipeline is performing we tested the response to a
unit impulse.  By taking the inner product of the impulse response for each channel
with the template, we can gauge very accurately the loss in
\SNR\ due to the approximations we have made and any inadequacies in
our implementation.

@@

In this section we check the effect of the \SVD\ tolerance parameter defined
in~\cite{Cannon:2010p10398}.  By changing the tolerance of the reconstruction
of the physical waveforms one effects directly the mismatch that results from
the truncation of the orthogonal filter matrix.  The conditions presented here
are more complicated than in the original work~\cite{Cannon:2010p10398} due to
the inclusion of the time sliced templates and resampling.  Figure
\ref{fig:hist-svd-tolerance} demonstrates that it is possible to acheive
typical \SNR\ losses of $\ll1\%$.  However, we do notice that the mismatch
saturates with respect to the tolerance.  This could be the effect of the
resampling, or another \SNR\ loss that we did not model or expect.  However the
saturation is still an order of magnitude below our target mismatch of
$\sim$0.003.  We find that an \SVD\ tolerance of $\left(1-10^{-4}\right)$ is
adequate to acheive our target SNR loss.
%
\begin{figure}
	\begin{center}
		\subfloat[Mismatch versus \textsc{svd} tolerance]{\label{fig:hist-svd-tolerance}\includegraphics{bw.pdf}}
		\subfloat[Mismatch versus $N^\shortuparrow$]{\label{fig:hist-interpolate}\includegraphics{bw_resample.pdf}}
		\caption{Box-and-whisker plot of mismatch between nominal
template bank and \lloid\ measured impulse respones.  The upper and lower boundaries of
the boxes show the upper and lower quartiles; the lines in the center denote the medians.
The whiskers represent the minimum and maximum mismatch over all templates.  In 
\subref{fig:hist-svd-tolerance} the interpolation filter length is held fixed
at $N^\shortuparrow = 192$, while the \SVD\ tolerance is varied from
$\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.  In \subref{fig:hist-interpolate}, the \SVD\ tolerance is fixed at $\left(1-10^{-6}\right)$ while $N^\shortuparrow$ is varied from 8 to 192 coefficients.}
	\end{center}
\end{figure}

@@

We are now prepared to offer the estimated computational cost of filtering this
sub-bank of templates compared to other methods.  We used the results of the
previous sub sections to set the \SVD\ tolerance to $1-10^{-4}$  and the
interpolation filter length to 16.  We find that the \lloid\ method provides
comparable performance to the standard \FD\ approach but with no intrinsic
latency.  Table \ref{table:flops} lists the costs for filtering this sub-bank.
%
\Table{\label{table:flops}Computational cost in \flops\ of the \TD\ method, the \FD\ method, and \lloid\ for the bank of $2 \times 657$ templates sampled at 4096 Hz for chirp masses $1.1955 \leqslant \mathcal{M} / M_\odot \leqslant 1.2045$.  A block size of $D = 2 N$ is used for the \fft{}s in the \FD\ method.  Both the \FD\ method and \lloid\ are five orders of magnitude faster than the conventional \TD\ method.  However, the \FD\ method has a latency of over half of an hour, whereas \lloid\ has no latency at all.}
\br
Method & \flops\ & latency (s) \\
\mr
Time-domain (\TD) convolution & $2.4\times10^{13}$ & 0 \\
Frequency-domain (\FD) convolution & $2.6\times10^8$ & 2201 \\
\lloid\ (this work) & $4.7\times10^8$ & 0 \\
\br
\end{tabular}
\end{indented}
\end{table}

>>>>

We examine the performance of the \lloid\ algorithm on a small region of
compact binary parameter space centered on typical NS-NS masses.  We begin by
constructing a template bank that spans component masses between
1~and~3~$M_\odot$ using a simulated Advanced LIGO noise curve~\cite{lal}.  Then
we create sub banks based on chirp mass slices in that parameter space.
Figure \ref{fig:tmpltbank} illustrates this procedure.
\begin{figure}[htbp]
	\begin{center}
		\includegraphics{figures/tmpltbank.png}
		\caption{\label{fig:tmpltbank}Placement of template parameters used in this paper.  The template bank consists of 98544 templates with component masses $m_1$, $m_2$, between 1~and~3~$M_\odot$.  We shall design a filter bank to search for a small subset of these template parameters with chirp masses $\mathcal M$ between 1.1955~and~1.2045~$M_\odot$.}
	\end{center}
\end{figure}
The result is that we obtain 657 templates with chirpmasses between 1.1955~and~1.2045~$M_\odot$.  With this
sub-bank we were able to construct an efficient time slice decomposition that consisted of 11 time slices
with sample rates between 32 and 4096 Hz shown in table \ref{tab:time_slices}.
\begin{table}
\begin{indented}
\caption{\label{tab:time_slices} Filter design for these 657 templates.  From left to right, this table shows the sample rate, time interval, number of samples, and number of orthogonal templates for each time slice.  The \SVD{} tolerance is varied from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.}
\item[]\begin{tabular}{rr@{,}lc*{6}{r}}
\br
\multicolumn{4}{c}{} &\centre{6}{$\log_{10}$ (1 - \textsc{svd} tolerance)} \\
\ns
\multicolumn{4}{c}{} &\crule{6} \\
$f^s$ (Hz) & $[t^s$&$t^{s+1})$ (s) & $N^s$ & -1 & -2 & -3 & -4 & -5 & -6 \\
\mr
4096 & [0&0.5) & 2048 & 1 & 4 & 6 & 8 & 10 & 14 \\
512 & [0.5&4.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 16 \\
256 & [4.5&12.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 15 \\
128 & [12.5&76.5) & 8192 & 6 & 20 & 25 & 28 & 30 & 32 \\
64 & [76.5&140.5) & 4096 & 1 & 8 & 15 & 18 & 20 & 22 \\
64 & [140.5&268.5) & 8192 & 1 & 7 & 21 & 25 & 28 & 30 \\
64 & [268.5&396.5) & 8192 & 1 & 1 & 15 & 20 & 23 & 25 \\
32 & [396.5&460.5) & 2048 & 1 & 1 & 3 & 9 & 12 & 14 \\
32 & [460.5&588.5) & 4096 & 1 & 1 & 7 & 16 & 18 & 21 \\
32 & [588.5&844.5) & 8192 & 1 & 1 & 8 & 26 & 30 & 33 \\
32 & [844.5&1100.5) & 8192 & 1 & 1 & 1 & 12 & 20 & 23 \\
\br
\end{tabular}
\end{indented}
\end{table}
We use this template bank and decomposition for the remainder of this section.

@@

It is also within reason that some \SNR\ loss could arise from other
suboptimalities of implementation in our test pipeline.  To gauge
accurately how well the pipeline is performing we tested the response to a
unit impulse.  By taking the inner product of the impulse response for each channel
with the template, we can gauge very accurately the loss in
\SNR\ due to the approximations we have made and any inadequacies in
our implementation.

@@

In this section we check the effect of the \SVD\ tolerance parameter defined
in~\cite{Cannon:2010p10398}.  By changing the tolerance of the reconstruction
of the physical waveforms one effects directly the mismatch that results from
the truncation of the orthogonal filter matrix.  The conditions presented here
are more complicated than in the original work~\cite{Cannon:2010p10398} due to
the inclusion of the time sliced templates and resampling.  Figure~%
\ref{fig:hist-svd-tolerance} demonstrates that it is possible to acheive
typical \SNR\ losses of $\ll1\%$.  However, we do notice that the mismatch
saturates with respect to the tolerance.  This could be the effect of the
resampling, or another \SNR\ loss that we did not model or expect.  However the
saturation is still an order of magnitude below our target mismatch of
$\sim$0.003.  We find that an \SVD\ tolerance of $\left(1-10^{-4}\right)$ is
adequate to acheive our target \SNR{} loss.
%
\begin{figure}
	\begin{center}
		\subfloat[Mismatch versus \textsc{svd} tolerance]{\label{fig:hist-svd-tolerance}\includegraphics{bw.pdf}}
		\subfloat[Mismatch versus $N^\shortuparrow$]{\label{fig:hist-interpolate}\includegraphics{bw_resample.pdf}}
		\caption{Box-and-whisker plot of mismatch between nominal
template bank and \lloid\ measured impulse respones.  The upper and lower boundaries of
the boxes show the upper and lower quartiles; the lines in the center denote the medians.
The whiskers represent the minimum and maximum mismatch over all templates.  In 
\subref{fig:hist-svd-tolerance} the interpolation filter length is held fixed
at $N^\shortuparrow = 192$, while the \SVD\ tolerance is varied from
$\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.  In \subref{fig:hist-interpolate}, the \SVD\ tolerance is fixed at $\left(1-10^{-6}\right)$ while $N^\shortuparrow$ is varied from 8 to 192 coefficients.}
	\end{center}
\end{figure}

@@

We are now prepared to offer the estimated computational cost of filtering this
sub-bank of templates compared to other methods.  We used the results of the
previous subsections to set the \SVD\ tolerance to $1-10^{-4}$  and the
interpolation filter length to 16. Table \ref{table:flops} lists the costs for filtering this sub-bank. Both the \FD\ method and \lloid\ are five orders of magnitude faster than the conventional \TD\ method.  However, the \FD\ method has a latency of over half of an hour, whereas \lloid\ has no latency at all.
%
\Table{\label{table:flops}Computational cost in \flops\ of the \TD\ method, the \FD\ method, and \lloid\ for the bank of $2 \times 657$ templates sampled at 4096 Hz for chirp masses $1.1955 \leqslant \mathcal{M} / M_\odot \leqslant 1.2045$.  A block size of $D = 2 N$ is used for the \fft{}s in the \FD\ method.}
\br
Method & \flops\ & Latency (s) \\
\mr
Time-domain (\TD) convolution & $2.4\times10^{13}$ & 0 \\
Frequency-domain (\FD) convolution & $2.6\times10^8$ & 2201 \\
\lloid\ (this work) & $4.7\times10^8$ & 0 \\
\br
\end{tabular}
\end{indented}
\end{table}
